---
title: "Social Media Engagement Analysis"
author: "Lizbeth Pacheco"
date: "2026-05-04"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```


# Ingest the Dataset
```{r}
socialMedia <- read_excel("socialMediaEngagement.xlsx")
head(socialMedia)
```


### After looking over the dataset, the main focuses on this analysis are the following:
**Which social media platform has the highest engagement?**
**Does a higher sentiment score lead to higher engagement?**
**Which emotion type sees the lowest engagement?**




# Data Cleaning

**Checking the dataset structure**

```{r}
str(socialMedia)
```

**Checking for missing values**

```{r}
colSums(is.na(socialMedia))
```

```{r}
sum(duplicated(socialMedia))
```

**Prior to the analysis beginning, the dataset was checked for missing and duplicated values. All but one variable saw 0 missing values. *Mentions* has 3,941 missing values, but since it won't affect the analysis, there will be no change. There was 0 duplicates in the data.**




# Data Manipulation

**Average Engagement by Platform**
```{r}
platformEngagement <- group_by(socialMedia, platform)
summarise(platformEngagement,
          avgEngagement = mean(engagement_rate, na.rm = TRUE) * 100)
```
**Facebook seems to have the highest engagement rates out of all the platforms.**


**Positive Sentiment Posts with High Engagement Rates**
```{r}
highPositive <- filter(socialMedia,
       sentiment_label == "Positive" &
         engagement_rate >= 0.75)
```
```{r}
head(highPositive)
```
```{r}
nrow(highPositive)
```
**There are 244 posts that meet both of those filters.**


**Emotion Type with Lowest Engagement**

```{r}
emotionEngagement <- group_by(socialMedia, emotion_type)
summarise(emotionEngagement,
          avgEngagement = mean(engagement_rate, na.rm = TRUE) * 100) %>%
  arrange(avgEngagement)
```
**Posts that are associated with happy emotions seem to receive the lowest engagement.**




# Statistical Analysis

### **Descriptive Statistics**

**Mean**
```{r}
mean(socialMedia$engagement_rate) * 100
```
The mean shows the average level of engagement across all social media posts in this dataset is about 27%.

**Median**
```{r}
median(socialMedia$likes_count)
```
The median shows that the middle number of likes received across all social media posts in the dataset is 2,496.

**Minimum and Maximum**
```{r}
min(socialMedia$likes_count)
```
```{r}
max(socialMedia$likes_count)
```
These two show us the minimum and maximum likes received on posts within the dataset.


### **Hypothesis Testing**
Do positive sentiment posts receive higher engagement than negative sentiment posts?

```{r}
# Excluding 'Neutral' Posts
sentimentTest <- filter(socialMedia,
                   sentiment_label != "Neutral")
```

**Hypotheses**
Null Hypothesis (H~0~): There is no difference in engagement rates between positive and negative posts.
Alternative Hypothesis (H~1~): There is a difference in engagement rates between positive and negative posts.

### **t-Test**
```{r}
t.test(engagement_rate ~ sentiment_label,
       data = sentimentTest)
```
With the p-value being greater than 0.05, we fail to reject the null hypothesis. There is an insignificant amount of evidence to show that engagement rates differ between positive and negative posts.

### **Linear Regression**
Does sentiment score predict engagement rate?

```{r}
engagement_lm <- lm(engagement_rate ~ sentiment_score,
                    data = socialMedia)

summary(engagement_lm)
```
The model shows that the sentiment score cannot predict engagement rate. The p-value also shows a weak connection, sentiment scores does not affect engagement rate.




# Data Visualizations
**Platform Engagement**

```{r}
# Average engagement by platform
platformEngagement <- socialMedia %>%
  group_by(platform) %>%
  summarise(avgEngagement = mean(engagement_rate, na.rm = TRUE))
```

```{r}
ggplot(platformEngagement, aes(x = platform, y = avgEngagement)) +
  geom_col(fill = "palevioletred3") +
  labs(
    title = "Average Engagement Rate by Platform",
    x = "Platform",
    y = "Average Engagement Rate"
  ) +
  theme_minimal()
```

**Sentiment Score vs Engagement Rate**

```{r}
ggplot(socialMedia, aes(x = sentiment_score, y = engagement_rate)) +
  geom_point(color = "palevioletred", alpha = 0.5) +
  geom_smooth(method = "lm", color = "violetred3") +
  labs(
    title = "Relationship Between Sentiment Score and Engagement Rate",
    x = "Sentiment Score",
    y = "Engagement Rate"
  ) +
  theme_minimal()
```

**Engagement Rate by Emotion Types**

```{r}
# Average engagement rate by emotion type
emotionEngagement <- socialMedia %>%
  group_by(emotion_type) %>%
  summarise(avgEngagement = mean(engagement_rate, na.rm = TRUE))
```


```{r}
ggplot(emotionEngagement, aes(x = emotion_type, y = avgEngagement)) +
  geom_col(fill = "palevioletred3") +
  labs(
    title = "Average Engagement Rate by Emotion Type",
    x = "Emotion Type",
    y = "Average Engagement Rate"
  ) +
  theme_minimal()
```

